Sushi monster

### 2025-04-01 inst/jss_hyper3_real.Rnw
tuna types are maguro,toro,tekka

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
Equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then the first bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR_jss}.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and
\pkg{hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this is not
easily represented as a generalized linear model using the techniques
of~\citet{turner2020}.  Relatedlikelihood functions for drawn matches
were considered using the reified entity technique
of~\citet{hankin2020_jss} in which a drawn match was considered to be
won by a fictitious entity (a ``draw monster'') who possessed his own
BT strength.  However, the likelihood function of
table~\ref{windrawloss} has a number of desirable features not
present in the reified entity approach of~\citet{hankin2020_jss}.
Firstly, if we consider two players of identical strengths and nullify
the home ground effect (that is, $\lambda=1$) then the probability of
a draw is $\frac{2D}{2+2D}$; the odds of a draw is simply $D$,
regardless of team strength.  Secondly, considering only won (that is,
not score-drawn) games, the probability of a home win is
$\frac{\lambda}{1+\lambda}$, that is, the odds of a home win is simply
$\lambda$, independently of the player strength. Such arguments give
direct operational significance to $\lambda$ and $D$.

@@

<<testlambdazero, cache = TRUE, include = FALSE>>=
jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo = FALSE>>=
plot(log(s), L - max(L), type = 'b',
     xlab = expression(log(lambda)), ylab = 'Log-likelihood')
abline(v = 0, col = 'gray')
abline(h = c(0, -2), col = 'gray')
segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
\end{center}
\end{figure}

>>>>

The most general likelihood function would be
proportional to
%
\begin{equation}\label{mostgeneral}
  \prod_{i=1}^N
  \sum_{j=1}^n
 \left(\lambda_{ij}p_j\right)^{w_i}
\end{equation}
%
where the $p_j$ are traditional Bradley-Terry strengths, $p_j\geqslant
0$, $\sum p_j=1$, and the $\lambda_{ij}\geqslant 0$ are multiplicative
factors that account for ephemeral or temporary changes in competitive
ability.  The data $w_1,\ldots w_N$ correspond to the number of
``wins'' and number of matches, as per $w_{12},w_{34}$ in
Equation~\ref{like1234}.  In practice, almost all the $\lambda_{ij}$
are zero.  To fix ideas, consider the data presented in
Table~\ref{oldfirm}.  Ignoring draws for the moment, we have a
likelihood proportional to
%
\begin{table}[t!]
\centering
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{wins}} &
\multicolumn{4}{c@{}}{\small{Plays away, loses}}\\
\cmidrule(l){2-5}
&   Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers      &   NA &    81  &    69   &      15\\
   Celtic       &   76 &    NA  &    61   &      15\\
   Falkirk      &   17 &    17  &    NA   &       6\\
   Livingston   &    2 &     2  &     3   &      NA\\
\end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{draws}} &
\multicolumn{4}{c@{}}{\small{Plays away, draws}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &       NA  &   44 &     14   &       2\\
   Celtic      &       47  &   NA &     10   &       3\\
   Falkirk     &       12  &   19 &     NA   &       4\\
   Livingston  &       4   &   4  &     4    &      NA\\
   \end{tabular}
\\
\rule{0mm}{16mm}
\\
\begin{tabular}{@{} *{5}{c} @{}}
\headercell{\small{Plays at home,}\\ \small{loses}} &
\multicolumn{4}{c@{}}{\small{Plays away, wins}}\\
\cmidrule(l){2-5}
&  Rangers     &    Celtic & Falkirk & Livingston\\
\midrule
   Rangers     &   NA  &  42  &    2  &       0\\
   Celtic      &   46  &  NA  &    8  &       0\\
   Falkirk     &   57  &  43  &   NA  &       8\\
   Livingston  &   12  &  11  &    9  &      NA\\
  \end{tabular}
\caption{The \label{oldfirm} dominance of the Old Firm: 769 matches
  between Rangers, Celtic, Falkirk and Livingston arranged by home
  wins, draws, and home losses.  Data from {\tt stats.football.co.uk}}
\end{table}
%
\begin{equation}\label{firstfewRCFL}
  \left(\frac{\lambda p_R}{\lambda p_R + p_C}\right)^{81}\cdot
  \left(\frac{\lambda p_R}{\lambda p_R + p_F}\right)^{69}\cdots
  \left(\frac{\lambda p_L}{\lambda p_L + p_F}\right)^{3}\cdots
  \left(\frac{p_C}{p_C + \lambda p_R}\right)^{42}\cdots
    \left(\frac{p_F}{p_F + \lambda p_L}\right)^{9}
\end{equation}
%
where the home-ground advantage is represented by~$\lambda$.  In
package idiom, a term such as $\lambda p_R + p_C$ is represented as a
named numeric vector.  For example, if the home strength were $\lambda
= 1.8$ then one bracket would be {\tt c(Rangers=1.8, Celtic=1)}
with an associated power of $-81$.  The computational methodology is a
generalization of that used in~\cite{hankin2024_hyper3_jss} in which the
multiplicative factors correspond to number of clones.  However,
in~\cite{hankin2024_hyper3_jss} the factors are known non-negative
integers, and here they are, in general, unknown non-negative real
numbers.  The \class{hyper3} class enforces \pkg{disordR}
discipline~\citep{hankin2022_disordR_jss}.

@@

About 22\% of matches in Table~\ref{oldfirm} end in a draw and
\pkg{hyper2} formalism may be used to assess this.  Consider the
likelihood function described in Table~\ref{windrawloss}; this is not
easily represented as a generalized linear model using the techniques
of~\citet{turner2020}.  Related likelihood functions for drawn matches
were considered using the reified entity technique
of~\citet{hankin2020_jss} in which a drawn match was considered to be
won by a fictitious entity (a ``draw monster'') who possessed his own
BT strength.  However, the likelihood function of
table~\ref{windrawloss} has a number of desirable features not
present in the reified entity approach of~\citet{hankin2020_jss}.
Firstly, if we consider two players of identical strengths and nullify
the home ground effect (that is, $\lambda=1$) then the probability of
a draw is $\frac{2D}{2+2D}$; the odds of a draw is simply $D$,
regardless of team strength.  Secondly, considering only won (that is,
not score-drawn) games, the probability of a home win is
$\frac{\lambda}{1+\lambda}$, that is, the odds of a home win is simply
$\lambda$, independently of the player strength. Such arguments give
direct operational significance to $\lambda$ and $D$.

@@

<<testlambdazero, cache = TRUE, eval = TRUE, echo = FALSE>>=
jj <- optimize(f = function(D){f(c(lambda=1, D))}, interval = c(0.1, 0.4), maximum = TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

\begin{figure}[htbp]
  \begin{center}
<<plotlikesushi, echo = FALSE, fig = TRUE>>=
plot(log(s), L - max(L), type = 'b',
     xlab = expression(log(lambda)), ylab = 'Log-likelihood')
abline(v = 0, col = 'gray')
abline(h = c(0, -2), col = 'gray')
segments(x0 = log(sushimaxlambda), y0 =  0.1, y1 = -1.0, col = 'red' )
segments(x0 = log(sushi_lower    ), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(sushi_upper    ), y0 = -2.2, y1 = -1.8, col = 'blue')
@ 
\caption{A\label{profilesupportsushi} profile support function for the
  exchangeability term $\lambda$ in the case of sushi preferences.
  Vertical gray line shows $H_0\colon\lambda =1$, horizontal lines
  show the two-units-of-support limits with the credible interval of
  $(\Sexpr{round(sushi_lower,2)},\Sexpr{round(sushi_upper,2)})$,
  marked in blue.  Red bar shows evaluate at
  $\lambda\simeq\Sexpr{round(sushimaxlambda,2)}$}
\end{center}
\end{figure}

### 2025-04-01 inst/sushi.Rmd
soft link for hyper2.bib
tuna types are maguro,toro,tekka

---
title: "Sushi monster"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib  # copy vignettes/hyper2.bib to inst/
---

@@

```{r}
sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
a[] <- sushi_types[a]
noquote(a)
sushi_eq_classes <-
    c("ebi"  = 1, "anago" = 2, "maguro" = 1, "ika" = 1, "uni" = 1,
      "sake" = 1, "tamago" = 1, "toro" = 2, "tekka" = 2,
      "kappa" = 1)
```

@@

```{r calc_sushi}
sushi <- make_sushi_H(a, lambda = jj$maximum)
```

```{r calc_sushimaxp}
(sushi_maxp <- maxp(sushi))
```

>>>>

---
title: "Sushi monster"
author: "R. K. S. Hankin"
output: bookdown::html_document2
bibliography: hyper2.bib
---

@@

```{r}
sushi_types <- c("ebi", "anago", "maguro", "ika", "uni", "sake", "tamago", "toro", "tekka", "kappa")
a[] <- sushi_types[a]
noquote(a)
sushi_eq_classes <-
    c("ebi"  = 1, "anago" = 1, "maguro" = 2, "ika" = 1, "uni" = 1,
      "sake" = 1, "tamago" = 1, "toro" = 2, "tekka" = 2,
      "kappa" = 1)
```

@@

```{r calc_sushi, cache=TRUE}
sushi <- make_sushi_H(a, lambda = jj$maximum)
```

```{r calc_sushimaxp, cache=TRUE}
(sushi_maxp <- maxp(sushi))
```
